### 2015-02-26 Manuscript/Missing_mammals-draft.tex
Manuscript update (wrapped up results)

%But missing data in living taxa problem
However, the total evidence method requires a lot of data, both molecular and morphological for both living and fossil taxa. Entire sections of this data can sometimes be difficult or impossible to collect for every taxa present in the analysis. For example, fossil have really rarely molecular data available and morphological characters are rarely collected when molecular data is available (e.g. \cite{O'Leary08022013} \textit{vs.} \cite{meredithimpacts2011}). Therefore, the ability of this method to recover correct topology is expected to decrease with missing data and especially when few morphological data overlaps between the living and the fossil taxa \cite{GuillermeCooper}. %Owwwww the sneaky self cite ;-).
The effect of missing data is most important on topology when too few living taxa have available morphological data \cite{GuillermeCooper}.
For example, if there is no morphological available for any living taxa of a clade, it is impossible to link a fossil to this clade because no morphological data will overlap between the fossil (regardless the amount of data available for the fossil) and the living taxa (that have no morphological data). This property of the total evidence can lead to important topological incongruities because the fossil will only be able to branch to a clade of living taxa that contains morphological data, even if in reality, the fossil clearly does not belong to this clade (see figure ~\ref{Figure_missing_data_problem}).

\begin{figure}[!htbp]
\centering
    \includegraphics[width=1\textwidth]{MissingDataFigure.pdf}
\caption{Example of topological errors due to missing morphological data in living taxa. If a phylogeny contains two clades, for example Proboscidea and Cetartiodactyla, with molecular data for both but only morphological data for Proboscidea. If an additional Cetartiodactyla fossil (with no molecular data) will be added to the phylogeny, it will erroneously branch with the Proboscidea clade instead of the Cetartiodactyla one.}
\label{Figure_missing_data_problem}
\end{figure}

%What we're doing
It is therefore crucial to understand the distribution of the available morphological data for living taxa in a clade before using a total evidence approach because missing living taxa with morphological data can lead to two topological artefacts: (1) the impossibility for a fossil to branch in the right clade if there is no morphological data available in this clade; and (2) the higher probability for a fossil to branch within a wrong clade that has more morphological data available for living taxa than the right clade.

In this study, we assess the level of available data in living mammals in order to highlight the two caveats described above.
%In this study we investigate the amount of living mammal taxa with available morphological data to detect the two potential caveats described above that could lead to topological errors when building total evidence mammal phylogenetic trees.
We extracted living mammals taxa with available morphological data from 256 phylogenetic matrices available online and measured the proportion of morphological data availability for each mammalian order (1). Additionally, in the mammalian orders where data was missing, we estimated the structure of the available data to detect if the available data was biases towards certain clades only or if it was randomly distributed (2).

@@

\section{Material and Methods}
\subsection{Matrices search}
To investigate the available living taxa with morphological data, we downloaded morphological matrices from three main public databases: morphobank (\texttt{http://www.morphobank.org/}), Graeme Lloyd's website (\texttt{http://graemetlloyd.com/}) and Ross Mounce's GitHub repository (\texttt{https://github.com/rossmounce}). We downloaded all the matrices containing any fossil or living mammal taxa from these data bases. Additionally we ran a thorough Google Scholar search for matrices that might not have been uploaded on the previously cited data bases. We downloaded the eventual additional morphological matrices from any of the 20 first papers matching with our selected key words and with any of the 35 taxonomic levels (see supplementary materials for detailed description of the procedure). We downloaded 256 % @@@ Number of matrices
matrices containing a total of 9411 % @@@ Number of living OTUs
operational taxonomic units (OTUs) from the combination of both searches (public repositories and Google Scholar).

We then transformed all the matrices to be in the same nexus format. We then standardised the taxonomic nomenclature by fixing invalid binomial inputs to match with the official taxonomic nomenclature rules (i.e. \textit{H. sapiens} was transformed in \textit{Homo sapiens}). We assigned each species as being either living or fossil using a taxonomic matching algorithm. We designated as living OTU all the OTUs that where either present in \cite{FritzTree} or \cite{wilson2005mammal}. We designated as fossil OTUs all the OTUs that where present in the Paleobiology database. For the OTUs neither labelled as living or fossil we tried to decompose the OTUs name (i.e. $Homo\_sapiens$ became $Homo$ and $sapiens$) and tried to match the to Wilson Reeders taxonomy at any taxonomic level (Family, Genus, $Genus\_species$, etc.). The matching OTUs where labelled as living and the ones still not matching where ignored and labelled as non-applicable (NA; see supplementary material for more details on the taxonomic matching algorithm).

\subsection{Data availability analysis}
\subsubsection{Number of characters threshold}
From all the 256 % @@@ Number of matrices
matrices, we selected only the ones that had at least 100 morphological characters. This arbitrary threshold of a minimal number of morphological characters was chosen to be in adequacy with \cite{GuillermeCooper} and \cite{harrisonamong-character2014}. Also this threshold avoids biases towards small matrices that could be either not informative (e.g. too few characters not allowing character overlap among matrices; see discussion) or made of non-applicable characters (e.g. antlers which are sexual dimorphic characters proper to a specific clade).

\subsubsection{Data availability}
To assess the data availability per mammal order, we calculated the percentage of OTUs with morphological data for three different taxonomic levels: Family, Genera and Species. We highlighted all the orders containing less than 25\% of living taxa with morphological data because their amount of missing data ($>$75\%) was higher than in \cite{GuillermeCooper} and therefore highly probable of suffering from the effect of missing data.

\subsubsection{Available data structure}
For the orders with no morphological data for all its OTUs at the three different taxonomic levels (Family, Genera and Species), we investigated the structure of the available data to test if it was either (i) randomly distributed, (ii) over-dispersed or (iii) clustered. To measure the structure of the available data we used classic community structure metric from the \texttt{picante} R package \cite{picante}. We compared the structure of the available data for each order to the structure of a potentially fully sampled order (i.e. only the OTUs with available morphological data \textit{versus} all the OTUs).
For each orders and taxonomic levels that presented OTUs with no available morphological data, we calculated the Net Relatedness Index (NRI) which quantifies the overall distribution of the data with negative values showing more dispersed data and positive values more clustered data than expected by the null model \cite{webb2002phylogenies}. We choose to present only the NRI values because they have been shown to be slightly less sensitive to the structure of the phylogeny (i.e. branch length and topology) \cite{NRI,journal.pone.0004390} but we also calculated the two other common phylogenetic structure indices: Faith's Phylogenetic Distance (PD) \cite{Faith19921} and the Nearest Taxon Index (NTI) \cite{webb2002phylogenies}. Both metrics are available in the Supplementary results.

@@

\subsection{Data availability}
We extracted 1422 living mammal OTUs from the 256 matrices with a minimum of 6 characters and a maximum of 4541. After removing all the matrices with less than 100 characters, the number of extracted living mammals OTUs was down to 815. 11/28 orders have less than 25\% of taxa with morphological data at a species level and 24/28 orders have less than 75\% taxa with available morphological data. At the Genus level however only 3/28 orders have less than 25\% of taxa with morphological data and 16/28 have less than 75\%. Finally, at the family level no order has less than 25\% taxa with available morphological data and only 5/28 have less than 75\% (table~\ref{Table_morpho_taxa_proportion}).

@@

\subsection{Available data structure}
Among the orders containing at least one OTU with no available morphological data, only two orders (Carnivora and Chiroptera) are significantly clustered both at the species and the genus level but not at the family level (table~\ref{Table_data_structure}).
\renewcommand\baselinestretch{1.2}\selectfont
\begin{center}
\input{Table_data_structure}
\end{center}
\renewcommand\baselinestretch{2}\selectfont

Two contrasted results are shown on figure ~\ref{Figure_example_coverageA} and ~\ref{Figure_example_coverageB} with randomly distributed data in Cetartiodactyla (figure ~\ref{Figure_example_coverageA}) and clustered available data in Carnivora (mainly Canida; figure ~\ref{Figure_example_coverageB}). 
\begin{figure}[!htbp]
\centering
    \includegraphics[width=1\textwidth]{example_coverageA.pdf}
\caption{Distribution of available morphological data across Cetartiodactyla. Edges are colored in grey when no morphological data is available or in red when data is available.}
\label{Figure_example_coverageA}
\end{figure}

@@

\section{Discussion}
%Data collection
We know we might have missed some data but we performed a thorough and repeatable search and still managed to collected 1422 OTUs which represents 38\% of all living mammals

@@

%\section{SOM}

\input{Supplementary/Supplementary}

>>>>

%But missing data in living taxa problem
However, the total evidence method requires a lot of data, both molecular and morphological for both living and fossil taxa. Entire sections of this data can sometimes be difficult or impossible to collect for every taxa present in the analysis. For example, fossil have really rarely molecular data available and morphological characters are rarely collected when molecular data is available (e.g. \cite{O'Leary08022013} \textit{vs.} \cite{meredithimpacts2011}). Therefore, the ability of this method to recover correct topology is expected to decrease with missing data and especially when few morphological data overlaps between the living and the fossil taxa \cite{GuillermeCooper}. %Owwwww the sneaky self cite ;-).
The effect of missing data is most important on topology when too few living taxa have available morphological data \cite{GuillermeCooper}.
For example, if there is no morphological available for any living taxa of an entire clade, it is impossible to link a fossil taxon to this clade because no morphological data will overlap between the fossil (regardless the amount of data available for the fossil) and the living taxa (that have no morphological data). This property of the total evidence method can lead to important topological incongruities because the fossil will only be able to branch to a clade of living taxa that contains morphological data, even if in reality, the fossil taxon does not belong to that clade (see figure ~\ref{Figure_missing_data_problem}).

\begin{figure}[!htbp]
\centering
    \includegraphics[width=1\textwidth]{MissingDataFigure.pdf}
\caption{Example of topological errors due to missing morphological data in living taxa. If a phylogeny contains two clades, for example Proboscidea and Cetartiodactyla, with molecular data for both (in grey) but only morphological data for Proboscidea only (in red). If an additional Cetartiodactyla fossil (with no molecular data) is added to the phylogeny, it will erroneously branch with the Proboscidea clade instead of the Cetartiodactyla one because no morphological data will overlap between the fossil Cetartiodactyla and the living ones.}
\label{Figure_missing_data_problem}
\end{figure}

%What we're doing
It is therefore crucial to understand the distribution of the available morphological data for living taxa in a clade before using a total evidence approach because missing living taxa with morphological data can lead to two topological artefacts: (1) the impossibility for a fossil to branch in the right clade if there is no morphological data available in this clade \cite{GuillermeCooper}; and (2) the higher probability for a fossil to branch within a wrong clade that has more morphological data available for living taxa than the right clade \cite{GuillermeCooper}.

In this study, we assess the level of available data in living mammals in order to highlight the two caveats described above.
%In this study we investigate the amount of living mammal taxa with available morphological data to detect the two potential caveats described above that could lead to topological errors when building total evidence mammal phylogenetic trees.
We extracted living mammals taxa with available morphological data from 256 phylogenetic matrices available online and measured the proportion of morphological data availability for each mammalian order. Additionally, in the mammalian orders where data was missing, we estimated the structure of the available data to detect if the available data was biases towards certain clades only or if it was randomly distributed using the Net Relatedness Index \cite{webb2002phylogenies}.

@@

%\newpage

\section{Material and Methods}
\subsection{Matrices search}
To investigate the available living taxa with morphological data, we downloaded morphological matrices from three main public databases: morphobank (\texttt{http://www.morphobank.org/}) \cite{morphobank}, Graeme Lloyd's website (\texttt{http://graemetlloyd.com/}) and Ross Mounce's GitHub repository (\texttt{https://github.com/rossmounce}). We downloaded all the matrices containing any fossil or living mammal taxa from these databases. Additionally we ran a thorough Google Scholar search for matrices that might not have been uploaded on the previously cited data bases. We downloaded the additional morphological characters matrices from the 20 first Google search results matching with our selected keywords and with any of the 35 taxonomic levels (see supplementary materials for detailed description of the procedure). We downloaded 256 % @@@ Number of matrices
matrices containing a total of 9411 % @@@ Number of living OTUs
operational taxonomic units (OTUs) from the combination of both searches (public repositories and Google Scholar). The list of matrices is available in the supplementary materials.

We then transformed all the matrices to be in the same nexus format. We also standardised the taxonomic nomenclature by fixing invalid binomial inputs to match with the official taxonomic nomenclature rules (i.e. \textit{H. sapiens} was transformed in \textit{Homo sapiens}). We assigned each species as being either living or fossil using a taxonomic matching algorithm. We designated as living OTU all the OTUs that where either present in \cite{FritzTree} or \cite{wilson2005mammal}. We designated as fossil OTUs all the OTUs that where present in the Paleobiology database. For the OTUs neither labelled as living or fossil we tried to decompose the OTUs name (i.e. \textit{Homo$\_$sapiens} became \textit{Homo} and \textit{sapiens}) and tried to match the first element with any taxonomic level (Family, Genus, Species, etc.) from \cite{wilson2005mammal}. The matching OTUs where labelled as living and the ones still not matching where ignored and labelled as non-applicable (NA; see supplementary material for more details on the taxonomic matching algorithm).

\subsection{Data availability analysis}
\subsubsection{Number of characters threshold}
From all the 256 % @@@ Number of matrices
matrices, we selected only the ones that displayed at least 100 morphological characters per OTU. This arbitrary threshold of a minimal number of morphological characters was chosen to be in adequacy with \cite{GuillermeCooper} and \cite{harrisonamong-character2014}. Also this threshold avoids biases towards small matrices that could be either not informative \cite{wagner2000} (e.g. too few characters not allowing character overlap among matrices) or made of non-applicable characters \cite{Brazeau2011} (e.g. antlers which are sexual dimorphic characters proper to a specific clade).

\subsubsection{Data availability}
To assess the data availability per mammal order, we calculated the percentage of OTUs with morphological data for three different taxonomic levels: Family, Genus and Species. We highlighted all the orders containing less than 25\% of living taxa with morphological data because their amount of missing data ($>$75\%) was higher than in \cite{GuillermeCooper} and therefore highly probable of having wrong topology due to the missing data. On the opposite, orders with up to 25\% of missing data (75\% of available data) were shown to have no significant effect on topology \cite{GuillermeCooper}. We therefore used this value as a threshold of "good" data sampling.

\subsubsection{Available data structure}
For the orders with no morphological data for all its OTUs at the three different taxonomic levels (Family, Genus and Species), we investigated the structure of the available data to test if it was either (i) randomly distributed, (ii) over-dispersed or (iii) clustered. To measure the structure of the available data we used classic community structure metric from the \texttt{picante} R package \cite{picante}. We compared the structure of the available data for each order to the structure of a potentially fully sampled order (i.e. only the OTUs with available morphological data \textit{versus} all the OTUs).
For each orders and taxonomic levels that presented OTUs with no available morphological data, we calculated the Net Relatedness Index (NRI) which quantifies the overall distribution of the data with negative values showing more dispersed data and positive values more clustered data than expected by the null model (random) \cite{webb2002phylogenies}. We choose to present only the NRI values because they have been shown to be slightly less sensitive to the structure of the phylogeny (i.e. branch length and topology) \cite{NRI,journal.pone.0004390} but we also calculated the two other common phylogenetic structure indices: Faith's Phylogenetic Distance (PD) \cite{Faith19921} and the Nearest Taxon Index (NTI) \cite{webb2002phylogenies}. Both metrics are available in the Supplementary results.

@@

\subsection{Data availability}
We extracted 1422 living mammal OTUs from the 256 matrices with a minimum of 6 characters and a maximum of 4541 per OTU. After removing all the matrices with less than 100 morphological characters, the number of extracted living mammals OTUs was down to 815. 11/28 orders have less than 25\% of taxa with morphological data at a species level and 24/28 orders have less than 75\% taxa with available morphological data. At the Genus level however only 3/28 orders have less than 25\% of taxa with morphological data and 16/28 have less than 75\%. Finally, at the family level no order has less than 25\% taxa with available morphological data and only 5/28 have less than 75\% (table~\ref{Table_morpho_taxa_proportion}).

@@

\subsection{Available data structure}
Among the orders containing at least one OTU with no available morphological data, only four orders are significantly clustered: Dirpotodontia, Carnivora and Chiroptera both at the species and the genus level and Rodentia at the species level only (table~\ref{Table_data_structure}).

\renewcommand\baselinestretch{1.2}\selectfont
\begin{center}
\input{Table_data_structure}
\end{center}
\renewcommand\baselinestretch{2}\selectfont

Two contrasted results are shown on figure ~\ref{Figure_example_coverageA} and ~\ref{Figure_example_coverageB} with randomly distributed data in Cetartiodactyla (figure ~\ref{Figure_example_coverageA}) and clustered available data in Carnivora (mainly Canidae; figure ~\ref{Figure_example_coverageB}).

\begin{figure}[!htbp]
\centering
    \includegraphics[width=1\textwidth]{example_coverageA.pdf}
\caption{Distribution of available morphological data across Cetartiodactyla. Edges are colored in grey when no morphological data is available or in red when data is available.}
\label{Figure_example_coverageA}
\end{figure}

@@

\section{Discussion}
Our results show that even though relations among living mammals are well known in a molecular framework (e.g. \cite{FritzTree,meredithimpacts2011,May-Collado-PeerJ}) only little morphological data is available. This is primarily due to the fact that that morphological studies are mainly focusing on fossil taxa rather than living taxa (e.g. \cite{O'Leary08022013,ni2013oldest}). However, data availability varies in its structure as well as across taxonomic levels. Higher taxonomic levels are always better sampled than lower ones (table~\ref{Table_morpho_taxa_proportion}) and within these taxonomic levels, the structure of available data is mostly random apart for four orders (table\ref{Table_data_structure}).

%Data collection
We know we might have missed some data but we performed a thorough and repeatable search and still managed to collected 1422 OTUs which represents 38\% of all living mammals

\begin{figure}[!htbp]
\centering
    \includegraphics[width=1\textwidth]{../andrew/citations.jpg}
\caption{ }
\label{bla}
\end{figure}

@@

%\section{SOM}
\newcommand{\beginsupplement}{%
    \setcounter{table}{0}
    \renewcommand{\thetable}{S\arabic{table}}%
    \setcounter{figure}{0}
    \renewcommand{\thefigure}{S\arabic{figure}}%
}
\beginsupplement
\input{Supplementary/Supplementary}

### 2015-02-26 Manuscript/Supplementary/Supplementary.tex
Manuscript update (wrapped up results)

\bibliographystyle{vancouver}
\bibliography{Supp_References}

\section{Data structure}
%Tables
%Figures

\section{Supplementary results}

>>>>

%\bibliographystyle{vancouver}
%\bibliography{Supp_References}

\section{Supplementary results}
The following section contains supplementary results to the main body: the available data structure using the NTI and the PD metric; the proportion of available data and the data structure for all the matrices (including the matrices with less than 100 characters); and phylogenetical representation of the data availability per order (excluding Cetartiodactyla and Carnivora, present in the main body).

\renewcommand\baselinestretch{1.2}\selectfont
\begin{center}
\input{Supplementary/Supp_data_structureNTI}
\end{center}
\renewcommand\baselinestretch{2}\selectfont

\renewcommand\baselinestretch{1.2}\selectfont
\begin{center}
\input{Supplementary/Supp_data_structurePD}
\end{center}
\renewcommand\baselinestretch{2}\selectfont

\renewcommand\baselinestretch{1.2}\selectfont
\begin{center}
\input{Supplementary/morpho_taxa_proportion_threshold1}
\end{center}
\renewcommand\baselinestretch{2}\selectfont

\renewcommand\baselinestretch{1.2}\selectfont
\begin{center}
\input{Supplementary/Supp_data_structureNRI_threshold1}
\end{center}
\renewcommand\baselinestretch{2}\selectfont

\renewcommand\baselinestretch{1.2}\selectfont
\begin{center}
\input{Supplementary/Supp_data_structureNTI_threshold1}
\end{center}
\renewcommand\baselinestretch{2}\selectfont

\renewcommand\baselinestretch{1.2}\selectfont
\begin{center}
\input{Supplementary/Supp_data_structurePD_threshold1}
\end{center}
\renewcommand\baselinestretch{2}\selectfont

\begin{figure}[!htbp]
\centering
    \includegraphics[width=1\textwidth]{Supplementary/Supp_figure_AFROSORICIDA.pdf}
\caption{Distribution of available morphological data across Afrosoricida. Edges are colored in grey when no morphological data is available or in red when data is available.}
\label{Supp_Figure_Phylo-Afrosoricida}
\end{figure}

\begin{figure}[!htbp]
\centering
    \includegraphics[width=1\textwidth]{Supplementary/Supp_figure_CHIROPTERA.pdf}
\caption{Distribution of available morphological data across Chiroptera. Edges are colored in grey when no morphological data is available or in red when data is available.}
\label{Supp_Figure_Phylo-Chiroptera}
\end{figure}

\begin{figure}[!htbp]
\centering
    \includegraphics[width=1\textwidth]{Supplementary/Supp_figure_CINGULATA.pdf}
\caption{Distribution of available morphological data across Cingulata. Edges are colored in grey when no morphological data is available or in red when data is available.}
\label{Supp_Figure_Phylo-Cingulata}
\end{figure}

\begin{figure}[!htbp]
\centering
    \includegraphics[width=1\textwidth]{Supplementary/Supp_figure_DASYUROMORPHIA.pdf}
\caption{Distribution of available morphological data across Dasyuromorphia. Edges are colored in grey when no morphological data is available or in red when data is available.}
\label{Supp_Figure_Phylo-Dasyuromorphia}
\end{figure}

\begin{figure}[!htbp]
\centering
    \includegraphics[width=1\textwidth]{Supplementary/Supp_figure_DIDELPHIMORPHIA.pdf}
\caption{Distribution of available morphological data across Didelphimorphia. Edges are colored in grey when no morphological data is available or in red when data is available.}
\label{Supp_Figure_Phylo-Didelphimorphia}
\end{figure}

\begin{figure}[!htbp]
\centering
    \includegraphics[width=1\textwidth]{Supplementary/Supp_figure_DIPROTODONTIA.pdf}
\caption{Distribution of available morphological data across Diprotodontia. Edges are colored in grey when no morphological data is available or in red when data is available.}
\label{Supp_Figure_Phylo-Diprotodontia}
\end{figure}

\begin{figure}[!htbp]
\centering
    \includegraphics[width=1\textwidth]{Supplementary/Supp_figure_ERINACEOMORPHA.pdf}
\caption{Distribution of available morphological data across Erinaceomorpha. Edges are colored in grey when no morphological data is available or in red when data is available.}
\label{Supp_Figure_Phylo-Erinaceomorpha}
\end{figure}

\begin{figure}[!htbp]
\centering
    \includegraphics[width=1\textwidth]{Supplementary/Supp_figure_PILOSA.pdf}
\caption{Distribution of available morphological data across Pilosa. Edges are colored in grey when no morphological data is available or in red when data is available.}
\label{Supp_Figure_Phylo-Pilosa}
\end{figure}

\begin{figure}[!htbp]
\centering
    \includegraphics[width=1\textwidth]{Supplementary/Supp_figure_PRIMATES.pdf}
\caption{Distribution of available morphological data across Primates. Edges are colored in grey when no morphological data is available or in red when data is available.}
\label{Supp_Figure_Phylo-Primates}
\end{figure}

\begin{figure}[!htbp]
\centering
    \includegraphics[width=1\textwidth]{Supplementary/Supp_figure_RODENTIA.pdf}
\caption{Distribution of available morphological data across Rodentia. Edges are colored in grey when no morphological data is available or in red when data is available.}
\label{Supp_Figure_Phylo-Rodentia}
\end{figure}

\begin{figure}[!htbp]
\centering
    \includegraphics[width=1\textwidth]{Supplementary/Supp_figure_SCANDENTIA.pdf}
\caption{Distribution of available morphological data across Scandentia. Edges are colored in grey when no morphological data is available or in red when data is available.}
\label{Supp_Figure_Phylo-Scandentia}
\end{figure}

\begin{figure}[!htbp]
\centering
    \includegraphics[width=1\textwidth]{Supplementary/Supp_figure_SORICOMORPHA.pdf}
\caption{Distribution of available morphological data across Soricomorpha. Edges are colored in grey when no morphological data is available or in red when data is available.}
\label{Supp_Figure_Phylo-Soricomorpha}
\end{figure}
